```{R}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(stats)
library(tidyverse)
library('fmsb')
library(grid)
library(broom)
library(gridExtra)
library(ComplexUpset)
library(UpSetR)
library(ggforestplot)
library(viridis)
library(colorblindr)
library(patchwork)
library(igraph)
library(ggraph)
library(tidyverse)
```

# Part A: CAZY 

#### 1: Start by reading in the dbcan results. Make df with file names
```{r}

chosen=c("Exophiala_alcalophila", "Exophiala_crusticola", "Exophiala_lecanii-corni", "Exophiala_mesophila", "Exophiala_spinifera", "Exophiala_viscosa", "Exophiala_xenobiotica")

cazy.files=data.frame(names=str_replace(list.files("CAZY/"), ".tsv", "")) %>%
  filter(names %in% chosen) %>%
  mutate(loc=paste("CAZY/", names, ".tsv", sep="")) %>%
  mutate(source=as.character(1:length(loc)), Genome=as.factor(c("Exophiala_alcalophila", "Exophiala_crusticola", "Exophiala_lecanii-corni", "Exophiala_mesophila", "Exophiala_spinifera", "Exophiala_viscosa", "Exophiala_xenobiotica")))
head(cazy.files)

```
#### 2: make a single dataframe with the dbcan hits for each genome and add a column label for each genome
```{r}
cazy.raw=cazy.files$loc %>% map_dfr(read.delim, .id="source", header=F) %>%
  left_join(cazy.files %>% select(source, Genome)) %>%
  select(-source)
head(cazy.raw)
```

#### 3: Make a dataframe with each CAZY annotation, the genome it is in, and the accession of the CAZY domain
```{r}

cazy=cazy.raw %>%
  select(V3, V1, Genome) %>%
  rename(Accession=`V3`, CAZY=`V1`) %>%
  mutate(CAZY=str_replace_all(CAZY, ".hmm", ""))
head(cazy)

```

#### 4: count number of times a unique accession appears in each genome
```{r}
cazy.counts=cazy %>%
  group_by(Genome, CAZY) %>%
  summarize(Count=n_distinct(Accession)) %>%
  group_by(Genome) %>%
  mutate(n=sum(Count)) %>%
  group_by(CAZY) %>%
  mutate(n_Genomes=length(unique(Genome)), Genomes=toString(unique(Genome)))
head(cazy.counts)

```
#### 5: get counts of unique cazy domains for each genome, and also a column with list of all the cazy domains in each genome for the upset plot
```{r}
cazy.composition=cazy.counts %>%
  group_by(Genome) %>%
  summarize(n=sum(n_Genomes), CAZYs=list(unique(CAZY)))
  head(cazy.composition)
```

#### 6: make the upset plot for cazy. Note that it was not very interesting so not including in paper
```{r}
cazy.dat=cazy.counts %>%
  filter(n_Genomes>1) %>%
  select(-n_Genomes) 
  head(cazy.dat)
cazy.phylo=levels(as.factor(cazy.dat$Genome))[c(1, 2, 5, 6, 7, 3, 4)]

cazy.upset.dat=cazy.composition$CAZYs
names(cazy.upset.dat)=cazy.composition$Genome
head(cazy.upset.dat)

cazy.uplt=UpSetR::upset(fromList(cazy.upset.dat), sets=cazy.phylo, mb.ratio = c(0.55, 0.45), order.by = "freq", keep.order = TRUE)

cazy.uplt
```
##### Note that the result was not very interesting so not including in paper as there is only one unique domain for E. crusticola

#### 7: Check the only unique domain for E crusticola
```{r}
cazy.unique=cazy.counts %>%
  filter(n_Genomes==1) %>%
  group_by(Genome) %>%
  summarize(n=n_distinct(CAZY), cazy=list(unique(CAZY)))
  head(cazy.unique)

print(cazy.unique[cazy.unique$Genome=="Exophiala_crusticola",]$cazy)
```
##### Note: expected to find more unique cazy domains... This one might be involved in EPS production but not really sure. Probably it's a function of copy

#### 8: function for enrichment of cazy

##### Note: Removed this part
##### Note: nothing was significant for E crusticola compared to genus members, have to compare to other outgroups
##### Note: I can try doing fisher's exact test between only a couple genomes to see if that changes things, for example E crusticola and E dermatidis. 
##### I can also use the function from the original script once I include outgroups since it seemed to work for comparing all genes from all genomes

#### 9: Make plot for fold change in cazy domain count

```{r}
# get median counts for genes across the genus
cazy.meta=cazy.dat %>%
  select(CAZY, Genome, Count) %>%
  rename(Domain="CAZY") %>%
  group_by(Domain) %>%
  mutate(avg=median(Count), Fold=ifelse((Count+avg)>0, Count/avg, 0),
         Direction=ifelse(Fold>1 | Fold==Inf, "Up", ifelse(Fold==1 | Fold==0, ifelse(Count<avg, "Down", "Equal"), "Down"))) %>%
  mutate(Fold=ifelse(Direction=="Down", 1/Fold, Fold))
  head(cazy.meta, 30)


# print only values above 2 fold in genome E crusticola
Ecrusticola <- cazy.meta %>% filter(Genome=="Exophiala_crusticola" & Fold>1)

# get counts for all the genomes for the domains in CTH in d, then filter out the domains in CTH
other_data <- cazy.meta %>% filter(Domain %in% Ecrusticola$Domain) # %>% filter(Genome != "CTH")

cazy_plot <- ggplot(data = other_data, aes(y = reorder(Domain, Fold), x = Genome, color = Direction, size = Fold)) +
  geom_point(alpha = 0.8) +  # Points for other genomes
  geom_stripes(inherit.aes = FALSE, aes(y = reorder(Domain, Fold)), odd = "#33333333", even = "#00000000") +  # Add striped background
  theme_minimal() +  # Base theme
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 15),  # Rotate x-axis text
        axis.text.y = element_text(size = 15),  # Y-axis text size
        legend.position = "right") +  # Legend position
  ylab("Domain") +  # Y-axis label
  scale_color_manual(values = c("Up" = "#004488", "Equal" = "#DDAA33", "Down" = "#BB5566")) +  # Custom colors for Direction
  labs(color = "Direction\nvs. median", size = "Significant\nDifferences") +  # Labels for legend
  coord_flip() +  # Flip coordinates
  guides(color = guide_legend(override.aes = list(size = 3), order = 1))  # Adjust legend appearance

# Show the plot
print(cazy_plot)

```
##### 10: make multifacet plot with bar
```{R} 
# Calculate difference from genus mean
diff_data <- Ecrusticola %>%
  filter(Genome == "Exophiala_crusticola") %>%
  mutate(Diff = Count - avg)  # ΔCount = Count - avg

# Create the bar plot
cazy_bar_plot <- ggplot(diff_data, aes(x = reorder(Domain, Diff), y = Diff, fill = Diff > 0)) +
  geom_col() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "none") +
  scale_fill_manual(values = c("TRUE" = "#004488", "FALSE" = "#BB5566")) +  # Green for increase, red for decrease
  labs(x = "Pfam Domain", y = "ΔCount (E. crusticola - Genus Mean)") +
  coord_flip()

# Show the plot

#ggsave(p, filename = "exophiala_crusticola_CAZY_plot_multipannel.png", width = 20, height = 10)

```

# Part B: Pfam domains 
#### 1: read in signalP results for crusticola
```{R}
sigp <- read.table("CAZY/Exophiala_crusticola.run_dbcan/signalp.out", quote="\"", comment.char="")
head(sigp)

```
#### 2: add column to denote these accessions are secreted
```{R}
#Combining noTm/TM results
sigp2 = sigp %>%
  select(V1, V12) %>%
  rename(`Accession`="V1", `Method`="V12") %>%
  mutate(Prediction="Secreted") %>%
  distinct()

```
#### 3: Get locations for pfam annotations
```{R}
chosen=c("Exophiala_alcalophila", "Exophiala_crusticola", "Exophiala_lecanii-corni", "Exophiala_mesophila", "Exophiala_sideris", "Exophiala_spinifera", "Exophiala_viscosa", "Exophiala_xenobiotica")

pfam.files=data.frame(names=str_replace(list.files("Pfam/"), ".domtbl", "")) %>%
  filter(names %in% chosen) %>%
  mutate(loc=paste("Pfam/", names, ".domtbl", sep="")) %>%
  mutate(source=as.character(1:length(loc)), Genome=as.factor(c("Exophiala_alcalophila", "Exophiala_crusticola", "Exophiala_lecanii-corni", "Exophiala_mesophila", "Exophiala_sideris", "Exophiala_spinifera", "Exophiala_viscosa", "Exophiala_xenobiotica")))
head(pfam.files)
```

#### 4: read in raw files for pfam annotations
```{r}
pfam.raw=pfam.files$loc %>% map_dfr(read_table2, .id="source", comment="#", col_names=F) %>%
  left_join(pfam.files %>% select(source, Genome)) %>%
  select(-source)
head(pfam.raw)
# check against the dom table to make sure it was read in properly
```
#### 5: Simplify dataframe so that it only has the domain accessions and the genome it came from
```{r}
pfam=pfam.raw %>%
  select(X4, X1, Genome) %>%
  rename(Accession=`X1`, Pfam=`X4`)
head(pfam, 20)
```
#### 6: build dataframe with counts of secreted/not secreted predictions for each domain
```{r}
sigp.pfam=pfam %>%
  left_join(sigp2, by = "Accession") %>%
  mutate(Prediction=replace_na(Prediction, "Not Secreted")) %>%
  group_by(Pfam, Prediction) %>%
  summarize(Accessions_n=length(unique(Accession)), Accessions=toString(unique(Accession))) %>%
  group_by(Pfam) %>%
  mutate(Prediction_n=n_distinct(Prediction), Predictions=toString(unique(Prediction), sep="\n")) %>%
  pivot_wider(names_from="Prediction", values_from=c("Accessions_n", "Accessions")) %>%
  rename(Secreted="Accessions_n_Secreted", `Not Secreted`="Accessions_n_Not Secreted") %>%
  mutate(Secreted=replace_na(Secreted, 0),
         `Not Secreted`=replace_na(`Not Secreted`, 0)) %>%
  group_by(Pfam) %>%
  mutate(Secretion=ifelse(Prediction_n==2, "Both", Predictions)) %>%
  rename("Domain"="Pfam")

sigp.pfam2 = sigp.pfam %>%
select(Domain, Secretion)
head(sigp.pfam2)
```

#### 7: df with Pfam, genome source, counts

```{r}
pfam.counts.acc=pfam %>%
  group_by(Genome, Pfam) %>%
  summarize(Count=n_distinct(Accession)) %>%
  group_by(Genome) %>%
  mutate(n=sum(Count)) %>%
  group_by(Pfam) %>%
  mutate(n_Genomes=n_distinct(Genome), Genomes=toString(unique(Genome))) %>%
  mutate(Method="Accessions")
  head(pfam.counts.acc)
```
#### 8: df with all domains listed for each genome in a single column where genomes are rows

```{r}
pfam.composition=pfam.counts.acc %>%
  group_by(Genome) %>%
  summarize(n=sum(n_Genomes), pfams=list(unique(Pfam)))
  head(pfam.composition)
```
#### 9: df with all unique pfam domains for each genome

```{r}
pfam.unique=pfam.counts.acc %>%
  filter(n_Genomes==1) %>%
  group_by(Genome) %>%
  summarize(n=n_distinct(Pfam), pfam=list(unique(Pfam)))
  head(pfam.unique)

```
#### 10: Make upset plot 

```{r}
pfam.dat=pfam.counts.acc %>%
  filter(n_Genomes>1) %>%
  select(-n_Genomes) 
  head(pfam.dat)


pfam.phylo=levels(as.factor(pfam.dat$Genome))[c(1, 2, 5, 6, 7, 3, 4)]

pfam.upset.dat=pfam.composition$pfams
names(pfam.upset.dat)=pfam.composition$Genome

pfam.uplt=UpSetR::upset(fromList(pfam.upset.dat), sets=pfam.phylo, mb.ratio = c(0.55, 0.45), order.by = "freq", keep.order = TRUE)

pfam.uplt
grid.text("Pfam UpSet Plot",x = 0.65, y=0.95, gp=gpar(fontsize=20))
```

#### 11: enrichment analysis of pfam domains across genomes from Jason's paper
```{R}
enrich = function(dat, var='DOMAIN', group="Genome", thresh=0.01){
  out=data.frame()
  vars=unique(dat[[var]])
  len=length(vars)
  for(i in 1:len){
    sub=vars[i]
    #print(sub)
    tmp=dat[dat[[var]]==sub,]
    Ct=tmp$Count
    num=tmp$n
    names(Ct)=tmp[[group]]
    test=broom::tidy(pairwise.fisher.test(Ct, num, p.adjust.method="bonferroni"))
    if(dim(test)[1]>0){
      test2=data.frame(test) %>%
        filter(p.value<thresh) %>%
        mutate(Domain=sub)}else(test2=NA)
    out=rbind(out, test2)
  }
  return(out)
} #Enrichment function

pfam.res=enrich(pfam.dat, var="Pfam", group="Genome")

```
#### 12: Plot enrichment plot using network analysis

```{R}
# Convert the edge list into an igraph object
edges <- pfam.res[, c("group1", "group2", "Domain")]  # Include Domain column
graph <- graph_from_data_frame(edges, directed = FALSE)

# Define a custom color palette
pallette <- c("#F5793A", "#A95AA1", "#85C0F9", "#0F2080")

# Prepare the edges data with the -log10(p-value) for edge thickness
E(graph)$log_p <- -log10(pfam.res$p.value)
E(graph)$domain <- pfam.res$Domain  # Add Domain information to edges

network <- ggraph(graph, layout = "fr") +  # Using "fr" layout for force-directed layout
  geom_edge_link(aes(edge_alpha = log_p, edge_width = log_p, color = domain)) +  # Edge color based on Domain
  geom_node_point(size = 12, color = "darkgrey") +  # Node size and color
  geom_node_text(aes(label = name), size = 5, color = "black", repel = TRUE) +  # Node labels
  theme_void() +  # Remove grid lines and background
  scale_edge_color_manual(values = pallette) +  # Use custom color palette
  scale_edge_alpha(range = c(0.5, 1)) +  # Adjust transparency for more vivid edges
  scale_edge_width(range = c(1, 5))  # Adjust edge thickness for better visibility

```

#### 13: determine fold change for pfam domains in Exophiala_crusticola
```{R}
pfam.meta=pfam.dat %>%
  select(Pfam, Genome, Count) %>%
  rename(Domain="Pfam") %>%
  group_by(Domain) %>%
  mutate(avg=median(Count), Fold=ifelse((Count+avg)>0, Count/avg, 0),
         Direction=ifelse(Fold>1 | Fold==Inf, "Up", ifelse(Fold==1 | Fold==0, ifelse(Count<avg, "Down", "Equal"), "Down"))) %>%
  mutate(Fold=ifelse(Direction=="Down", 1/Fold, Fold))
```
#### 14: plot the fold change across genomes for crusticola and others for pfam vs median

```{r}

# print only values above 2 fold in genome E crusticola
Ecrusticola_pfam <- pfam.meta %>% filter(Genome=="Exophiala_crusticola" & Fold>2)

# get counts for all the genomes for the domains in CTH in d, then filter out the domains in CTH
other_data_pfam <- pfam.meta %>% filter(Domain %in% Ecrusticola_pfam$Domain) # %>% filter(Genome != "CTH")

pfam_plot <- ggplot(data = other_data_pfam, aes(y = reorder(Domain, Fold), x = Genome, color = Direction, size = Fold)) +
  geom_point(alpha = 0.8) +  # Points for other genomes
  geom_stripes(inherit.aes = FALSE, aes(y = reorder(Domain, Fold)), odd = "#33333333", even = "#00000000") +  # Add striped background
  theme_minimal() +  # Base theme
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 15),  # Rotate x-axis text
        axis.text.y = element_text(size = 15),  # Y-axis text size
        legend.position = "right") +  # Legend position
  ylab("Domain") +  # Y-axis label
  scale_color_manual(values = c("Up" = "#004488", "Equal" = "#DDAA33", "Down" = "#BB5566")) +  # Custom colors for Direction
  labs(color = "Direction\nvs. median", size = "Significant\nDifferences") +  # Labels for legend
  coord_flip() +  # Flip coordinates
  guides(color = guide_legend(override.aes = list(size = 3), order = 1))  # Adjust legend appearance

# Show the plot
print(pfam_plot)

```
##### 10: make bar plot for Exophiala crusticola pfam domains to show count differences
```{R} 
# Calculate difference from genus mean
diff_data_pfam <- Ecrusticola_pfam %>%
  filter(Genome == "Exophiala_crusticola") %>%
  mutate(Diff = Count - avg)  # ΔCount = Count - avg

# Create the bar plot
pfam_bar_plot <- ggplot(diff_data_pfam, aes(x = reorder(Domain, Diff), y = Diff, fill = Diff > 0)) +
  geom_col() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "none") +
  scale_fill_manual(values = c("TRUE" = "#004488", "FALSE" = "#BB5566")) +  # Green for increase, red for decrease
  labs(x = "Pfam Domain", y = "ΔCount (E. crusticola - Genus Mean)") +
  coord_flip()

# Show the plot
pfam_bar_plot
```
##### 11: make bar plot for unique pfam domains in E crusticola 

```{R} 

# take list of unique Pfam domains for crusticola and convert to a dataframe where Pfam domains are rows
pfam.unique2=pfam.unique %>% filter(Genome=="Exophiala_crusticola") %>% select(pfam) %>% unnest(pfam)

# join the pfam domains with the counts from the pfam.counts.dom dataframe, extract only value from Count column
pfam.unique3=pfam.unique2 %>% left_join(pfam.counts.acc, by=c("pfam"="Pfam"))
print(pfam.unique3, n=100)

# Create a bar plot

unique_pfam <- ggplot(pfam.unique3, aes(x = reorder(pfam, Count), y = Count, fill = Genome)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +  # Flip coordinates for better visibility
  labs(title = "Unique Pfam Domains in Exophiala crusticola",
       x = "Protein Domains",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_fill_manual(values = c("#DDAA33"))
```

##### 12: make multipannel plot for E crusticola domains

```{R} 
multiplot <- (pfam_bar_plot | pfam_plot) / (cazy_bar_plot | cazy_plot) / (network | unique_pfam)

ggsave(multiplot, filename = "exophiala_crusticola_multipannel.png", width = 30, height = 40)
```

##### 13: make multipannel plot for E crusticola domains
```{R} 
library(ggplot2)
library(dplyr)

# Convert 'Down' values to negative fold change
pfam.meta_heat <- other_data_pfam %>%
  mutate(Fold = ifelse(Direction == "Down", -Fold, Fold))

# Create the heatmap
pfam_heatmap <- ggplot(data = pfam.meta_heat, aes(x = Genome, y = reorder(Domain, Fold), fill = Fold)) +
  geom_tile(color = "white") +  # Heatmap with white gridlines
  geom_text(aes(label = Count), color = "black", size = 5) +  # Add count values inside cells
  scale_fill_gradient2(low = "#2166AC", mid = "#DDAA33", high = "#B2182B", midpoint = 0) +  
  theme_minimal() +  
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 15),  
    axis.text.y = element_text(size = 15),
    legend.position = "right"
  ) +
  labs(
    x = "Genome",
    y = "Domain",
    fill = "Fold Change"
  )

# Show the plot
print(pfam_heatmap)

```

##### 13: make multipannel plot for E crusticola cazy domains

```{R} 

# Convert 'Down' values to negative fold change
cazy.meta_heat <- other_data %>%
  mutate(Fold = ifelse(Direction == "Down", -Fold, Fold))

# Create the heatmap
cazy_heatmap <- ggplot(data = cazy.meta_heat, aes(x = Genome, y = reorder(Domain, Fold), fill = Fold)) +
  geom_tile(color = "white") +  # Heatmap with white gridlines
  geom_text(aes(label = Count), color = "black", size = 5) +  # Add count values inside cells
  scale_fill_gradient2(low = "#2166AC", mid = "#DDAA33", high = "#B2182B", midpoint = 0) +  
  theme_minimal() +  
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 15),  
    axis.text.y = element_text(size = 15),
    legend.position = "right"
  ) +
  labs(
    x = "Genome",
    y = "Domain",
    fill = "Fold Change"
  )

# Show the plot
print(cazy_heatmap)
multiplot2 <- (pfam_heatmap | cazy_heatmap) + plot_annotation(tag_levels = 'A')
ggsave(multiplot2, filename = "exophiala_crusticola_heatmaps.png", width = 30, height = 15)

```